Stack Safety is a Security Property

### 2021-04-14 tex/main.tex

We propose formal criteria for stack safety, using the concepts of
language-based security: a combination of an integrity property (``each
caller's stack frame is held invariant by the callee'') and a confidentiality
property (``the callee's behavior is insensitive to differences in the
caller's state'').
% These properties are sensible even in the presence of
% abnormal control flow.\apt{Is this important enough to mention here?}
% BCP: I think not in the abstract, but maybe in the intro.
% We further apply\apt{what does this mean? we didn't
%  invent it} a straightforward notion of
% ``well-bracketed control flow'' that restricts control flow to the
% typical pattern of matched calls and returns.
%
We argue that these criteria match common intuitions by (1) showing that
they are violated by a variety of intuitively unsafe examples and (2)
using property-based random testing to check that appropriate variants of
the criteria are actually enforced by the stack-safety ``micro-policies''
proposed by~\citet{DBLP:conf/sp/RoesslerD18} (and {\em not} enforced by
broken variants).  Specifically, Roessler
and Dehon's ``eager'' micro-policy, which catches violations as early as
possible, enforces a basic ``stepwise'' versions of our confidentiality
and integrity properties. Their more performant ``lazy'' micro-policy
does not, but corresponds to weaker, more extensional ``observational'' variants
of the
properties.  While we do not rule out other enforcement mechanisms, our
definitions are strongly informed by micropolicy systems, especially
our attacker model. \BCP{I am still not clear in what way, exactly.}

@@

In applying these security concepts, we treat each caller as needing protection
from its callee, even as that callee may need protection from its callee in turn.
This differs from common practice in language-based security in that the
confidentiality property applies to many nested subtraces within a single execution.\sna{How novel is this nesting?}\bcp{I believe completely novel?}
{\em Integrity} says that a caller's local data is protected from modification by its callee,
motivated by the reasoning principle that local data persists across a call. Its
{\em stepwise} interpretation formalizes this very naturally: a caller's local data
is never modified during a call. {\em Confidentiality} says that a caller's data is protected
from being leaked into its callee, and also that the callee cannot see uninitialized data.
As a reasoning principle, confidentiality guarantees that the callee's
behavior is determined only by a predictable interface that does not include the caller's
local state or the uninitialized portion of the stack. This is formalized as a
{\em noninterference}-style property that, rather than comparing pairs of whole program traces,
compares a pair of subtraces for each call, nested within other calls. Once again
{\em stepwise} describes the granularity of enforcement: at every step of a call,
changes in the state are independent of the caller's data.

Together, these properties capture the rough intuition of LSE, but are somewhat stronger.
First, we model returns as returns to a valid return target; an invalid return
is treated as a mere jump within the callee, and does not grant access to the caller's
data. So, even without WBCF, our stack safety properties rule out many control flow attacks.
The attacker in such an instance is permitted to ``return'' so long as it avoids
violating the integrity and confidentiality of the caller. Second, confidentiality
contains temporal protections that LSE does not implicitly address.
\apt{Maybe want to say a little more about this here, since it is a novelty wrt/ existing work?}
\bcp{+1}\sna{TODO: Say a little more, but make sure it's accurate first.}

@@

A second challenge is that
enforcement mechanisms may weaken their protections in the name of efficiency,
while arguing that they still capture stack safety in a meaningful sense.
For example, \citet{DBLP:conf/sp/RoesslerD18} describe how to protect the stack
using tag-based \emph{micro-policies}~\citep{pump_oakland2015}.
Memory cells within each stack frame are tagged with the
identity of the function activation
that owns the frame, and access to those locations is then gated
according to the activation that is currently executing. Roessler and DeHon
explore alternative versions of this basic idea, with different ways of identifying
functions and deciding exactly which accesses are allowed.
These vary in their efficiency and in the spatial and temporal precision of their
security guarantees.
The \emph{Depth Isolation} policy prevents a callee from writing outside its frame
at all, which is simple and intuitively correct, but it must manually clear the tags
on the frame on each return, which is very expensive. Conversely, the
\emph{Lazy Tagging and Clearing} policy improves
performance by assigning locations to the active function when they are written, which
permits  ``illicit'' writes to inaccessible stack frames so long as the
written locations are not subsequently read by their ``real'' owners. Intuitively,
this form of laziness is justified if any dangerous writes will
ultimately be detected before they can affect program behavior.

But what does ``affecting program behavior'' really mean? This micro-policy
calls for a second characterization of stack safety in terms of the externally
observable behavior of a stack-safe system. Such a characterization will have the
additional benefit of being more {\em extensional}, concerned primarily with the
external behavior of the system. In contract, our stepwise properties are {\em intensional},
focusing on internal details that may not actually matter to overall security.

To formalize this intuition, we develop an \emph{observational integrity} property which
characterizes enforcement mechanisms that detect violations between the point when
they occur and the point when they make an observable impact. Intuitively this property
can be phrased as ``no visible reads after dangerous writes.''  We prove (in Coq) that stepwise
integrity implies observational integrity.
\sna{There is a bit of massaging old proofs to do here, and I don't recall if we
  established that this was important.\apt{medium importance. it is obvious that
    observational integerity allows behaviors that stepwise does not, but
    maybe not obvious that the converse is not true too.}}
We additionally sketch an associated {\em observational confidentiality} property
makes confidentiality extensional in the same way, although we do not know of a real
enforcement mechanism that weakens confidentiality in this way.
\apt{why bother mentioning it then?}
\sna{If a relatively extensional characterization is a worthwhile contribution
  in its own right, then it's worth mentioning that we have one. If not, then indeed,
  we shouldn't mention it -- or get too excited about extensionality in general, because
  lockstep isn't extensional.}
Together, these form an {\em observational stack safety} that is highly extensional.
We therefore argue that, despite its somewhat greater complexity,
it may be a better minimal specification for
stack safety enforcement mechanisms in general.

\BCP{This paragraph doesn't apply to the lazy properties, so maybe it should
be moved before they are introduced?}%
To illustrate the applicability of these properties, we perform randomized testing to
validate that \citeauthor{DBLP:conf/sp/RoesslerD18}'s Depth Isolation micro-policy enforces
them. Randomized testing is a relatively low-cost way to validate a potential enforcement
mechanism; it finds flaws in the enforcement fast enough to be integrated into the workflow
of enforcement development \citep{TestingNI:ICFP}. To demonstrate that the testing adequately
covers potentially erroneous programs, we test intentionally buggy policies alongside
the real one. If the testing can differentiate correct implementations from buggy ones,
it is an argument in favor of their correctness. \bcp{Hmm--don't understand
  that sentence}
%\apt{more to say?}\bcp{Yes---partly to underscore our claim that this is a
%  significant contribution, and partly to give the reader some context for
%  why this is part of our story.  E.g., ``One nice thing we can do with our
%  spec is to use it to evaluate the claims of existing or future proposals
%  for enforcement mechanisms.  One relatively inexpensive way to carry out
%  such an evaluation is to see whether property-based testing against this
%  spec can distinguish correct implementations of the mechanism from buggy
%  ones...''}
Of course, as our testing reveals, \citeauthor{DBLP:conf/sp/RoesslerD18}'s Lazy Tagging and
Clearing micro-policy does not enforce stepwise integrity.
Unfortunately the existing Lazy Tagging and Clearing
micro-policy does not quite enforce observable confidentiality either. %, as we show in
%testing~\apt{really??},
But formalizing the ideal property allows us to see how it falls short -- a specific
violation of the temporal aspect of confidentiality.
We propose  a variant of Lazy Tagging and Clearing that we believe \emph{does} enforce
observational integrity, at some performance cost.\apt{but we're not testing that, right?}
\rb{Not sure about the property part right now, the machine side of this is
  pretty much done}
\sna{It's aspirational so far, but there's no fundamental reason we can't test it. We may
  need a caveat about how it's harder to test?}\apt{As per email discussion,
  maybe we could have a property variant that \emph{does} match depth-based Lazy Tagging and
  Clearing, by loosening the temporal safety properties?}
\sna{I'm pretty sure for the main policy it's just lazy integrity, no confidentiality.
  There was a variant Nick and Andre gave that would restrict the leaks to only be between
  instances of the same function, so that's the same property under a weaker attacker model,
  where we don't protect a function from itself.}

Our initial description of these properties assumes a single simple stack with no sharing
between callers and callees: all parameters and return values are passed in registers.  Later,
we show how to refine it to allow passing of scalar stack data, granting a
callee access to the caller's data. We do not model more sophisticated sharing,
such as taking the address of a variable and treating it like a pointer. Doing so without
protecting the variable is not interesting, and protecting would be better suited to
a different kind of enforcement, such as a memory safety micro-policy or Cheri-style capability.
\bcp{This is all a bit mysterious} \sna{Any less mysterious now?}

\rb{So no pointer passing? Pass by reference is now discussed at the end of
  Section 3, however it looks like things get tricky rather quickly.}
\sna{The trick is that if a programmer is passing pointers, we can't presume their
  intent -- how long should we permit memory to be shared? But if the compiler is passing
  pointers, as in pass by reference, we have a clear sense of where sharing should stop.
  \apt{N.B.Doing pass-by-reference properly requires that the passed variable is accessible
    by all descendents, not just immediate callee, which is doable with depth tags but not
    obviously with per-activation tags }}

@@

\section{Set Up and Threat Model}
\label{sec:threat}

\newcommand*{\rsp}{\textsc{sp}}

Our properties are defined in terms of an abstract machine model that makes very few
assumptions about the concrete machine that instantiates it.  We assume a register
machine with a program counter (\(\PCname\)); other
details of the machine's ISA are unimportant. We assume that we know ahead of time
the layout of memory, including where code lives and which addresses are within
the bounds of the stack.\bcp{This implies no run-time code generation,
  right?  This should probably be called out.}

\BCP{What should we say about the heap?}

@@

\sna{This is awkward.}
This division of policy state from machine state is helpful when reasoning about
enforcement mechanisms that keep a significant amount of data separate from the
primary execution data to inform security. Obviously tag-based systems do this,
with tags and extra policy state separated from and untouchable by application code.
Another significant example is the shadow stack, where a second stack
mirrors the real one, and a return may only occur if both stacks agree on the
target address. Typically the shadow stack is isolated using a separate protection
mechanism.

@@

\paragraph*{Threat Model}

@@

>>>>

We propose formal criteria for stack safety, using the concepts of
language-based security: a combination of an integrity property (``each
caller's stack frame is held invariant by the callee'') and a confidentiality
property (``the callee's behavior is insensitive to differences in the
caller's state'').
% These properties are sensible even in the presence of
% abnormal control flow.\apt{Is this important enough to mention here?}
% BCP: I think not in the abstract, but maybe in the intro.
% We further apply\apt{what does this mean? we didn't
%  invent it} a straightforward notion of
% ``well-bracketed control flow'' that restricts control flow to the
% typical pattern of matched calls and returns.
%
We argue that these criteria match common intuitions by (1) showing that
they are violated by a variety of intuitively unsafe examples and (2)
using property-based random testing to check that appropriate variants of
the criteria are actually enforced by the stack-safety ``micro-policies''
proposed by~\citet{DBLP:conf/sp/RoesslerD18} (and {\em not} enforced by
broken variants).  Specifically, Roessler
and Dehon's ``eager'' micro-policy, which catches violations as early as
possible, enforces a basic ``stepwise'' versions of our confidentiality
and integrity properties. Their more performant ``lazy'' micro-policy
does not, but corresponds to weaker, more extensional ``observational'' variants
of the
properties.  While we do not rule out other enforcement mechanisms, our
definitions are strongly informed by micropolicy systems, especially
our attacker model, which is stronger than other enforcement mechanisms
can enable.

@@

In applying these security concepts, we treat each caller as needing protection
from its callee, even as that callee may need protection from its callee in turn.
This is novel in that the
confidentiality property applies to many nested subtraces within a single execution,
as opposed to traces of entire executions.
{\em Integrity} says that a caller's local data is protected from modification by its callee,
motivated by the reasoning principle that local data persists across a call. Its
{\em stepwise} interpretation formalizes this very naturally: a caller's local data
is never modified during a call. {\em Confidentiality} says that a caller's data is protected
from being leaked into its callee, and also that the callee cannot see uninitialized data.
As a reasoning principle, confidentiality guarantees that the callee's
behavior is determined only by a predictable interface that does not include the caller's
local state or the uninitialized portion of the stack. This is formalized as a
{\em noninterference}-style property that, rather than comparing pairs of whole program traces,
compares a pair of subtraces for each call, nested within other calls. Once again
{\em stepwise} describes the granularity of enforcement: at every step of a call,
changes in the state are independent of the caller's data.

Together, these properties capture the rough intuition of LSE, but are somewhat stronger.
First, we model returns as returns to a valid return target; an invalid return
is treated as a mere jump within the callee, and does not grant access to the caller's
data. So, even without WBCF, our stack safety properties rule out many control flow attacks.
The attacker in such an instance is permitted to ``return'' so long as it avoids
violating the integrity and confidentiality of the caller. Second, confidentiality
contains temporal protections that LSE does not necessarily address. The intuition
for LSE is that it protects the caller's stack frame, not necessarily memory that was
previously allocated but now outside of the caller's frame. (Cheri-capability-based
implementations of LSE {\em do} protect previously allocated data, typically by clearing
it, but because it is a potential vector for leaking capabilities into the caller's frame,
not as a goal of the property.)

To illustrate the applicability of these properties, we perform randomized testing to
validate that an existing enforcement mechanism does enforce them.
\citet{DBLP:conf/sp/RoesslerD18} describe how to protect the stack
using tag-based \emph{micro-policies}~\citep{pump_oakland2015}, but offer no formal properties.
We test against their {\em Depth Isolation} micro-policy, in which memory cells within each stack
frame are tagged with the identity of the function activation
that owns the frame, and access to those locations is then gated
according to the activation that is currently executing. This is simple and corresponds
to intuitive notions of stack safety, and we argue that by describing its behavior
we approach a reasonable stack-safety property that other micro-policies can be
validated against.

Randomized testing is a relatively low-cost way to validate a potential enforcement
mechanism; it finds flaws in the enforcement fast enough to be integrated into the workflow
of enforcement development \citep{TestingNI:ICFP}. To demonstrate that the testing adequately
covers potentially erroneous programs, we test intentionally buggy policies alongside
the real one. If we can catch known bugs, we gain confidence that we can catch unknown ones,
strengthening the validation of the enforcement mechanism.

@@

A second challenge is that enforcement mechanisms may weaken their protections
in the name of efficiency, while arguing that they still capture stack safety in a meaningful sense.
\citet{DBLP:conf/sp/RoesslerD18} explore alternative micro-policies, with different
ways of identifying functions and deciding exactly which accesses are allowed.
These vary in their efficiency and in the spatial and temporal precision of their
security guarantees.
The Depth Isolation policy prevents a callee from writing outside its frame
at all, which is simple and intuitively correct, but it must manually clear the tags
on the frame on each return, which is very expensive. Conversely, their
\emph{Lazy Tagging and Clearing} policy improves
performance by assigning locations to the active function when they are written, which
permits  ``illicit'' writes to inaccessible stack frames so long as the
written locations are not subsequently read by their ``real'' owners. This violates our
straightforward properties, but ntuitively,
this form of laziness is justified if any dangerous writes will
ultimately be detected before they can affect program behavior.

What does ``affecting program behavior'' really mean? This micro-policy
calls for a second characterization of stack safety in terms of the externally
observable behavior of a stack-safe system. Such a characterization will have the
additional benefit of being more {\em extensional}, concerned primarily with the
external behavior of the system. In contract, our stepwise properties are {\em intensional},
focusing on internal details that may not actually matter to overall security.

To formalize this intuition, we develop an \emph{observational integrity} property which
characterizes enforcement mechanisms that detect violations between the point when
they occur and the point when they make an observable impact. Intuitively this property
can be phrased as ``no visible reads after dangerous writes.''  We prove (in Coq) that stepwise
integrity implies observational integrity. \sna{TODO: check this.}

We additionally sketch an associated {\em observational confidentiality} property
makes confidentiality extensional in the same way, in order to form an
{\em observational stack safety} that is highly extensional.
We therefore argue that, despite its somewhat greater complexity,
it may be a better minimal specification for
stack safety enforcement mechanisms in general.

%\apt{more to say?}\bcp{Yes---partly to underscore our claim that this is a
%  significant contribution, and partly to give the reader some context for
%  why this is part of our story.  E.g., ``One nice thing we can do with our
%  spec is to use it to evaluate the claims of existing or future proposals
%  for enforcement mechanisms.  One relatively inexpensive way to carry out
%  such an evaluation is to see whether property-based testing against this
%  spec can distinguish correct implementations of the mechanism from buggy
%  ones...''}
As it turns out, \citeauthor{DBLP:conf/sp/RoesslerD18}'s Lazy Tagging and
Clearing micro-policy enforces observational integrity, but not confidentiality,
whether stepwise or observational.
But formalizing the ideal property allows us to see how it falls short -- a specific
violation of the temporal aspect of confidentiality occurs in corner cases where callees
can leak data between one another.
We propose  a variant of Lazy Tagging and Clearing that we believe \emph{does} enforce
confidentiality, albeit at some performance cost.

Our initial description of these properties assumes a single simple stack with no sharing
between callers and callees: all parameters and return values are passed in registers.  Later,
we show how to refine it to allow passing of scalar stack data, granting a
callee access to the caller's data. We do not model more sophisticated sharing,
such as taking the address of a variable and treating it like a pointer.

@@

\section{Setup}
\label{sec:setup}

Our properties are defined in terms of an abstract machine model that makes very few
assumptions about the concrete machine that instantiates it.  We assume a register
machine with a program counter (\(\PCname\)); other
details of the machine's ISA are unimportant. We assume that we know ahead of time
the layout of memory, including where code lives and which addresses are within
the bounds of the stack. We do not model a heap; programs might choose to use memory
that is neither code nor stack to store global variables, as a heap, etc., but our
properties do not protect it or assume any particular behavior. We also do not
model run-time code generation.

@@

Dividing policy state from machine state helps us reason about
enforcement mechanisms that keep a significant amount of data separate from the
primary execution data to inform security. Obviously tag-based systems do this,
with tags and extra policy state separated from and untouchable by application code.
Another significant example is the shadow stack, where a second stack
mirrors the real one, and a return may only occur if both stacks agree on the
target address. Typically the shadow stack is isolated using a separate protection
mechanism.

@@

\section{Threat Model}
\label{sec:threat}

@@

We will describe observable integrity in our most detailed setting. In the
presence

\definition Let \(\stackid\) be a stack identifier, \(\depth\) be any depth, and
\(f_\depth ~ \mach ~ (\_, \rtm, \_, \stackid)\) hold for \(|\rtm ~ \stackid| \geq \depth\).
For any reachable segment \(\MPCT \langle f_\depth \rangle\),
let \(\head(\MPCT) = \mach,\pol,(\domMap,\rtm,\yts,\stackid)\). Let
\(\components\) be the set of components \(\component\) such that
\(\domMap ~ \component = \instack{\stackid}{\sealed}\) or \(\instack{\stackid}{\passed{\depth,\depth'}}\)
where \(|\rtm ~ \stackid| < \depth\) or \(\depth' < |\rtm ~ \stackid|\) .
Then a system enjoys observable

Then take any \(\nach\) such that \(\mach \approx_\components \nach\), and
\(\nach \hookrightarrow \NPCT\langle f_\depth \rangle\). A system enjoys
{\em stepwise stack confidentiality} in the coroutine setting if for all
such \(\MPCT\) and \(\NPCT\), \(\MPCT \doteq \NPCT\).
